Learning CNN filters from user-drawn image makers for coconut-tree image classification

### 2020-06-08 main.tex
Updates from Overleaf

The design of CNN models without the expert as part of the training loop leaves several research questions unanswered: (1) How to choose the most reasonable model, in the sense that it provides acceptable effectiveness, being as simple and efficient as needed, for a given classification problem? (2) How to train that model from a reduced number of samples?  (3) Can the user explain the decisions of that model? (4) Can that model improve from human corrections? The first question requires to explore human knowledge about the classification technique and the problem of interest. The second one requires to reduce as much as possible the human effort to train a model. The third issue is related to human understanding, and it may explore data visualization techniques to explain the decisions of the model and perhaps to guide human decisions concerning the design of the model. The fourth question is also important during training, and it is related to human control over the process. They all lead to the importance of involving human experts during the machine learning process.

In this paper, we present some advances towards answering the questions (1) and (2). We also discuss future research related to the questions (3) and (4). We present a method to the design of simplified convolutional neural networks from markers drawn by a human expert on a few images to indicate the most discriminant regions among the classes. From those markers, our approach can learn the filters of each convolutional layer, such that the CNN can be created in a layer-by-layer fashion, with no need for backpropagation in order to build a representative feature space for the classification problem. Backpropagation is only used to train a bank of multi-layer perceptrons for local pattern classification.

We propose a new way of learning the filters of the convolutional layers of deep neural networks that do not depend on the optimization of an error function and, therefore, do not need a large number of training samples. The training process is done with user intervention, which adds markers to the training images to indicate relevant features to characterize each class. To learn the filters, we get a patch around each marker of a class and cluster them. The centroid of each cluster is a filter that can be used to identify features of this cluster. The set of centroids of clusters of all clusters is the convolutional layer filters. For the next layer, the patches are generated on the output of the previous one. To evaluate our method, we use a dataset of aerial images to detect coconut trees. We identified that, with few layers and few training samples, we managed to get competitive results with state-of-the-art models.

@@

Let $F$ be a filter with dimension $k \times k$, and let $\pi$ be the patch around pixel $(x,y)$ also with dimension $k \times k$. By the definition of convolution in deep learning, we can interpret this operation for a pixel $(x, y)$ as the dot product between the kernel vector $\vect(F)$ and the vector of a patch $\vect(\pi)$ around this pixel, that is, the convolution in $(x, y)$ is the same as $\vect(F) \cdot \vect(\pi)$. Thus, let $H$ be a hyperplane orthogonal to vector $\vect(F)$. Vectors similar to $\vect(F)$ must fall on the same side of the hyperplane as $\vect(F)$. Therefore, if vector $\vect(\pi)$ is on the same side of $H$ as $\vect(F)$, then the doct product between $\vect(\pi)$ and $\vect(F)$ is positive, and it is negative otherwise. Thus, the doct product between the centroid of a cluster and each of the vectors of that cluster must be positive, which implies that the centroid is a good filter to identify the pattern described by that cluster. Hence, if we throw away the negative values, applying convolution with centroids as filters must highlight regions that has a certain pattern that the network designer believes is important.

Since all the patches used to produce the filters are centered around the origin, as well as the images of $\D$, we have that the hyperplanes ortogonal to the filters contain the origin, therefor there is no need for bias.

Each convolutional layer is trained individually, one layer at a time. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max pooling operation to aggregate local information. We preserve the output dimensions of convolutional layers so that we can use the same markers used in the previous layer. The new patches are generated from the previeus convolutional layer output and the whole process is repeated to find the filters of the next one.

@@

\section{Experiments}
We use a dataset formed from patches of aerial images made by \cite{8899005} from imagery provided by WeRobotics. The imagery with a resolution of 8cm was captured by the World Bank's UAVs for Disaster Resilience Program, collaborating with WeRobotics and OpenAerialMap in the Kindom of Tonga in October 2017. The Humanitarian OpenStreetMap community provided the annotation. The dataset consists of 13587 patches of dimensions $90 \times 90$ of coconut (10268) and non-coconut (3319). The latter class has a large variability. The problem of interest is to classify the images in one of the two classes. We only report experiments with one dataset due to the space limitation.

@@

\begin{table}[!t]
  \begin{center}
  \begin{tabular}{|l|c|c|c|}
  \hline
   Method & Precision & Recall & F-score \\
  \hline\hline
    I-Ours & $0.856 \pm 0.014$ & $0.836 \pm 0.020$ & $0.842 \pm 0.018$\\
    II-Ours (fine tuned) & $0.86 \pm 0.01$ & $0.84 \pm 0.02$ & $0.84 \pm 0.02$\\
    III-Ours with SVM & $0.836 \pm 0.015 $ & $ 0.793 \pm 0.025$ & $ 0.804 \pm 0.022$\\
    IV-VGG & $0.84 \pm 0.01$ & $0.76 \pm 0.03$ & $0.77 \pm 0.02 $ \\
    V-VGG (fine tuned) & $0.84 \pm 0.01$ & $0.75 \pm 0.03$ & $0.77 \pm 0.02 $ \\
  \hline
  \end{tabular}
  \end{center}
  \caption{Results.}
  \label{tab:results}
\end{table}

>>>>

The design of CNN models without a human expert as part of the training loop leaves several questions unanswered: (1) How to choose the most reasonable model, with acceptable effectiveness, and as simple as needed for a given classification problem? (2) How to train that model from a reduced number of images?  (3) Can the user explain the decisions of that model? (4) Can that model improve from human corrections? The first question requires to explore human knowledge about the classification technique and the problem of interest. The second one requires to reduce as much as possible the human effort to train a CNN model. The third issue is related to human understanding. It may explore data visualization techniques to explain the model's decisions and guide human decisions concerning the design of the model. The fourth question is also essential during training, and it is related to human control over the process. They all lead to the importance of involving human experts (users) during the machine learning process.

In this paper, we present significant advances related to the questions (1) and (2). First, we conceptually divide a CNN into convolutional layers for feature extraction and fully connected layers for feature space reduction and classification. Each convolutional layer contains a filter bank, an activation function, and alternative operations (e.g., pooling, batch normalization).  As the number of convolutional layers increases (deeper is the model), higher is the number of annotated samples required to train the model by backpropagation.

We present a method that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, a few training images, and draws markers in regions that best discriminate the classes. We explain how the classes may appear in distinct clusters of a local feature space defined by patches as extracted from marker pixels at the input of each convolutional layer. Patch size (number of weights per filter) is also a user choice. The architecture of the network can be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}. The method uses clustering and defines the weights of the filters from the centroids of the clusters. Such filters must activate the discriminant regions in the output of the given convolutional layer, and the process repeats in a layer-by-layer fashion with usually no need for new markers.

We demonstrate the advantages of this new technique on binary image classification of coconut-tree images against a popular CNN model, VGG-16~\cite{simonyan2014very}. First, the user can directly interfere and verify the markers' effectiveness in the training process.  The strategy allows user control and a better understanding of CNN models. Second, very few images per class (e.g., less than five) are usually enough. Third, by eliminating the need for optimization of convolutional layers by backpropagation, the method can considerably reduce the number of annotated images to train the fully connected layers.  Fourth, the technique is application-independent, and so it might be useful in other image classification problems.

@@

Let $F$ be a filter with dimension $k \times k$, and let $\pi$ be the patch around pixel $(x,y)$ also with dimension $k \times k$. By the definition of convolution in deep learning, we can interpret this operation for a pixel $(x, y)$ as the dot product between the kernel vector $\vect(F)$ and the vector of a patch $\vect(\pi)$ around this pixel, that is, the convolution in $(x, y)$ is the same as $\vect(F) \cdot \vect(\pi)$. Thus, let $H$ be a hyperplane orthogonal to vector $\vect(F)$. Vectors similar to $\vect(F)$ must fall on the same side of the hyperplane as $\vect(F)$. Therefore, if vector $\vect(\pi)$ is on the same side of $H$ as $\vect(F)$, then the dot product between $\vect(\pi)$ and $\vect(F)$ is positive, and it is negative otherwise. Thus, the dot product between the centroid of a cluster and each of the vectors of that cluster must be positive, which implies that the centroid is a good filter to identify the pattern described by that cluster. Hence, if we throw away the negative values, applying convolution with centroids as filters must highlight regions with a certain pattern that the network designer believes is important.

Since all patches used to produce the filters are centered around the origin, as well as the images of $\D$, we have that the hyperplanes orthogonal to the filters contain the origin. Therefore, there is no need for bias.

Each convolutional layer is trained individually, one layer at a time. After the convolution operation, we apply the ReLU function to eliminate negative activations and the max pooling operation to aggregate local information. We preserve the convolutional layers' output dimensions so that we can use the same markers used in the previous layer. The new patches are generated from the previous convolutional layer output, and the whole process is repeated to find the filters of the next one.

@@

\section{Experiments}
We use a dataset formed from patches of aerial images made by \cite{8899005} from imagery provided by WeRobotics. The imagery with a resolution of 8cm was captured by the World Bank's UAVs for Disaster Resilience Program, collaborating with WeRobotics and OpenAerialMap in the Kingdom of Tonga in October 2017. The Humanitarian OpenStreetMap community provided the annotation. The dataset consists of 13587 patches of dimensions $90 \times 90$ of coconut (10268) and non-coconut (3319). The latter class has a large variability. The problem of interest is to classify the images in one of the two classes. We only report experiments with one dataset due to the space limitation.

@@

\begin{table}[!t]
  \begin{center}
  \begin{tabular}{|l|c|c|c|}
  \hline
   Method & Precision & Recall & F-score \\
  \hline\hline
    I-Ours & $0.856 \pm 0.014$ & $0.836 \pm 0.020$ & $0.842 \pm 0.018$\\
    II-Ours (B) & $0.851 \pm 0.018$ & $0.834 \pm 0.023$ & $0.839 \pm 0.021$\\
    III-Ours (SVM) & $0.836 \pm 0.015 $ & $ 0.793 \pm 0.025$ & $ 0.804 \pm 0.022$\\
    IV-VGG & $0.84 \pm 0.01$ & $0.76 \pm 0.03$ & $0.77 \pm 0.02 $ \\
    V-VGG (PT) & $0.84 \pm 0.01$ & $0.75 \pm 0.03$ & $0.77 \pm 0.02 $ \\
  \hline
  \end{tabular}
  \end{center}
  \caption{Results.}
  \label{tab:results}
\end{table}
